%% This is an example first chapter.  You should put chapter/appendix that you
%% write into a separate file, and add a line \include{yourfilename} to
%% main.tex, where `yourfilename.tex' is the name of the chapter/appendix file.
%% You can process specific files by typing their names in at the 
%% \files=
%% prompt when you run the file main.tex through LaTeX.
\chapter{Conclusion} \label{ch:conc}
The ability to run as many VMs as possible
on individual hosts is an important determinant
in the cost-savings achievable
by server and desktop virtualization.
In this context, VM boot storms represent a major -- and often daily -- 
obstacle for data centers that wish to sustain a high VM density
per host in all workloads.

This thesis proposed a novel
design idea, {\em silhouette execution},
as a possible strategy for
mitigating the prohibitively high
CPU overhead that results from concurrently
booting VMs. In theory, silhouette execution 
exploits potential overlap in distinct instruction streams
of different VMs and uses it to avoid repetitive
execution on a host.

In order to understand whether
there is sufficient overlap in the instruction streams of
concurrently booting VMs to merit an implementation of 
silhouette execution, we used dynamic instrumentation 
techniques to study user-space execution traces of 
a few Linux services. Using our traces,
we were able to statistically profile
the extent of nondeterminism in these programs,
and also collect a comprehensive list of 
the sources behind such nondeterminism.
Through our experiments, we found that a
lot of the nondeterminism in Linux services
arises from the numerous -- and sometimes obscure --
interactions between user-space programs
and the lower software layers. We also
found that most of the nondeterminism
is an artefact of synthetic differences
in program execution (e.g. variable ordering
or latency of I/O events, unpredictability
of process IDs) rather than 
semantic differences in execution.

We outlined a few design sketches
for silhouette execution in user-space,
and used simple simulation techniques to
mathematically model the effectiveness of 
silhouette execution for a few Linux services. In
our simulations, synthetic differences in program execution
caused silhouette execution to fail and in fact
increased the number of instructions
executed in user space by 6\% for 1000 VMs.
However, using our knowledge of nondeterminism
in programs, we were able to propose
techniques that can increase
the instruction overlap 
between various program instances (e.g.
signal, I/O alignment and process ID virtualization).
Based on the few Linux services we sampled, silhouette execution
theoretically achieves a $19\times$ reduction
in user-space CPU pressure for 1000 VMs,
which is an encouraging result. 
The improvement in host scalability
during boot storms may be somewhat larger
because reduced CPU pressure can allow
transparent page sharing to be more
effective.

\section {Future Work}
There are several ways we can expand on the 
results presented in this thesis: \newline

\noindent{\bf Improve Silhouette Execution Modeling}\newline
Currently, we only evaluate silhouette
execution in user-space, by studying
nondeterminism in a few Linux services. 
In the future, we hope to consolidate
our results by:
\begin{itemize}
\item Incorporating results from a greater 
  number and variety of Linux services and
  distributions.
\item Finding more sophisticated models 
  for each of the design sketches
  for silhouette execution. Currently,
  the models for execution overhead are linear and
  conservative and do not account for
  execution latencies.
\item Incorporate nondeterminism
and instruction traces from lower software
layers (e.g. the kernel or hypervisor)
for completeness.
\end{itemize} 
\noindent{{\bf Implementation}}\newline
We are currently working on a user-space
implementation of silhouette execution 
that uses dynamic instrumentation.
In the future, we hope
our endeavors will serve as a foundation
for a complete implementation of silhouette execution
that provably mitigates boot storms. \newline

